---
title: "Sling"
author: "Sedat Kacar"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# 1- LOAD LIBRARIES and Prepare Slingshot object

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(Seurat)
library(SingleCellExperiment)
library(ggplot2)
library(dplyr)

SedatFunctions::rds_function(8)

# Assuming you have a Seurat object named 'seurat_obj'
sce_obj <- as.SingleCellExperiment(Endo_sickle)

sce_obj <- sce_obj[rowSums(assay(sce_obj, "counts")) > 0, ]



```

# 2- CALCULATE TRAJECTORIES using slingshot from the UMAP coordinates.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(slingshot)
pto <- slingshot(reducedDim(sce_obj,'UMAP'), clusterLabels = sce_obj$Concise_new1,
start.clus = 'gCaps', end.clus = c('aCaps', 'PAECs', 'PVECs', 'gCapC'), spar = 1.1)
```


# 3- Visualizing Slingshot-Inferred Lineage Paths on UMAP Using Minimum Spanning Tree (MST)

```{r, message=FALSE, warning=FALSE, echo=FALSE}
slsMST <- slingMST(pto, as.df = TRUE)

scater::plotReducedDim(sce_obj, dimred = 'UMAP', colour_by = 'Concise_new1') +
  ggplot2::geom_point(data = slsMST, aes(x = umap_1, y = umap_2), size = 3, color = "grey15") +
  ggplot2::geom_path(data = dplyr::arrange(slsMST, Order),
                     ggplot2::aes(x = umap_1, y = umap_2, group = Lineage)) +
  ggplot2::labs(title = "Slingshot MST on UMAP", x = "UMAP 1", y = "UMAP 2")


```

# 4- VISUALIZE ORIGINAL UMAP

```{r, message=FALSE, warning=FALSE, echo=FALSE}

hvg_genes <- VariableFeatures(Endo_sickle)
rowData(sce_obj)$highly_variable <- rownames(sce_obj) %in% hvg_genes


sce_obj <- sce_obj[rowSums(assay(sce_obj, "counts")) > 0, ]

scater::plotReducedDim(sce_obj, 'UMAP', colour_by = 'Concise_new1', text_by = 'Concise_new1') + theme_bw()


```



# 5- Map the inferred trajectories (lineages) from the cluster-level graph into the actual reduced-dimensional space 
```{r, message=FALSE, warning=FALSE, echo=FALSE}
emb <- slingshot::embedCurves(pto, reducedDim(sce_obj,'UMAP'))
emb <- slingshot::slingCurves(emb)

g <- scater::plotReducedDim(sce_obj, 'UMAP', text_by = 'Concise_new1', colour_by='Concise_new1')
for (path in emb) {
embedded_curves.df <- data.frame(path$s[path$ord,])
g <- g + ggplot2::geom_path(data=embedded_curves.df, aes(x=umap_1, y=umap_2), linewidth=1.2)
}
print(g)

```





```{r, message=FALSE, warning=FALSE, echo=FALSE}
sce_obj <- scater::runUMAP( sce_obj, dimred = 'PCA',
BPPARAM=BiocParallel::MulticoreParam(6),
ncomponents = 3, name = 'UMAP_3D'
)

```






```{r, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}

#1
sce.sling <- slingshot::slingshot(sce_obj, reducedDim='PCA', clusterLabels = sce_obj$Concise_new1,
start.clus = 'gCaps', end.clus = c('aCaps', 'PAECs', 'PVECs', 'gCapC'), spar = 1.1)

#2
slsMST <- slingshot::slingMST(pto, as.df = TRUE)
scater::plotReducedDim(sce_obj, dimred='PCA', colour_by = 'Concise_new1') +
ggplot2::geom_point(data = slsMST, aes(umap_1, umap_2), size = 3, color = "grey15") +
ggplot2::geom_path(data = slsMST %>% arrange(Order),
ggplot2::aes(umap_1, umap_2, group = Lineage)) +
ggplot2::labs(title = "Slingshot MST on UMAP", x = "PCA 1", y = "PCA 2")

```

# 6- Interactive 3D UMAP with Slingshot Lineages Colored by Cluster and Pseudotime

```{r, message=FALSE, warning=FALSE, echo=FALSE, include=FALSE}
#3
embedded <- slingshot::embedCurves(sce.sling, "UMAP")
embedded <- slingshot::slingCurves(embedded)

#4
g <- scater::plotReducedDim(sce.sling, 'UMAP', text_by = 'Concise_new1', colour_by='Concise_new1')
for (path in embedded) {
embedded_curves.df <- data.frame(path$s[path$ord,])
g <- g + ggplot2::geom_path(data=embedded_curves.df, aes(x=umap_1, y=umap_2), linewidth=1.2)
}

print(g)

#5
sce.sling$Pseudotime <- rowMeans(slingshot::slingPseudotime(sce.sling), na.rm=TRUE)
#6
g <- scater::plotReducedDim(sce.sling, 'UMAP', text_by = 'Concise_new1', colour_by='Pseudotime')
for (path in embedded) {
embedded_curves.df <- data.frame(path$s[path$ord,])
g <- g + ggplot2::geom_path(data=embedded_curves.df, aes(x=umap_1, y=umap_2), linewidth=1.2)
}
print(g)
#7
embedded <- embedCurves(sce.sling, 'UMAP_3D')


```



# 7- 3D Visualization of Slingshot-Inferred Lineages Colored by Cell Clusters in UMAP Space
```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
library(plotly)

?reducedDim
plot.df <- as.data.frame(reducedDim( sce.sling, 'UMAP_3D' ) )
plot.df$cluster <- sce.sling[['Concise_new1']]
colnames(plot.df) <- c('UMAP1', 'UMAP2', 'UMAP3', 'cluster')
redDim <- 'UMAP'

p <- plot_ly( ) %>% # colors = trace.colors ) %>%
  add_trace( data = plot.df,
             x = ~UMAP1, y = ~UMAP2, z = ~UMAP3,
             color = ~cluster,
             type = 'scatter3d',
             mode = 'markers',
             marker = list( size = 3, opacity = 1 )
           )

for (i in 1:4) {
  embedded_curves.tmp <- slingCurves(embedded)[[i]] # only 1 path
  embedded_curves.tmp.df <- data.frame(embedded_curves.tmp$s[embedded_curves.tmp$ord,])
  p <- p %>% 
    add_trace( data = embedded_curves.tmp.df,
               x = ~UMAP1, y = ~UMAP2, z = ~UMAP3, 
               type = 'scatter3d', mode = "lines",
               name = paste0("Lineage ", i),
               line = list( width = 5, opacity = 1 )
             )
}

p %>% 
  layout( scene = list( xaxis = list(title = paste0( redDim, '1') ),
                        yaxis = list(title = paste0( redDim, '2') ),
                        zaxis = list(title = paste0( redDim, '3') )
                      )
        )

```

# 8-Interactive 3D Visualization of Slingshot Trajectories Colored by Pseudotime in UMAP Space

```{r, message=FALSE, warning=FALSE, echo=FALSE}
plot.df$Pseudotime <- sce.sling$Pseudotime

p <- plot_ly( ) %>% # colors = trace.colors ) %>%
  add_trace( data = plot.df,
             x = ~UMAP1, y = ~UMAP2, z = ~UMAP3,
             type = 'scatter3d',
             mode = 'markers',
             showlegend = FALSE,
             marker = list( color = ~Pseudotime,
                            size = 3,
                            colorscale = 'Viridis', # Viridis color scale
                            colorbar = list( # Add color scale bar,
                              title = 'Pseudotime',
                              x = 1.01,       # horizontal position
                              y = 0.4,        # vertical position
                              len = 0.8,      # length (0 to 1, relative to plot height)
                              thickness = 30  # width in pixels
                            ), 
                            opacity = 1
                          )
           )

for (i in 1:4) {
  embedded_curves.tmp <- slingCurves(embedded)[[i]] # only 1 path
  embedded_curves.tmp.df <- data.frame(embedded_curves.tmp$s[embedded_curves.tmp$ord,])
  p <- p %>% 
    add_trace( data = embedded_curves.tmp.df,
               x = ~UMAP1, y = ~UMAP2, z = ~UMAP3, 
               type = 'scatter3d', mode = "lines",
               name = paste0("Lineage ", i),
               line = list( width = 5, opacity = 1 )
             )
}

p %>% 
  layout( scene = list( xaxis = list(title = paste0( redDim, '1') ),
                        yaxis = list(title = paste0( redDim, '2') ),
                        zaxis = list(title = paste0( redDim, '3') )
                      )
        )

```

```{r, message=FALSE, warning=FALSE, echo=FALSE,include=FALSE}
sce.sling <- scater::logNormCounts(sce.sling, assay.type="logcounts")
sce.sling

assayNames(sce.sling)

```


```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(TSCAN)
# scran::testPseudotime(...)
pseudo <- TSCAN::testPseudotime( sce.sling,  
                          pseudotime=sce.sling$Pseudotime, 
                          BPPARAM = BiocParallel::MulticoreParam(6)
                        )
```

###  9-Filter by Selecting significant  genes

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)
# ensembldb::filter(., FDR < 0.05)
pseudo.sig <- pseudo %>%
  as.data.frame() %>%
  dplyr::filter( FDR < 0.05 ) %>%
  arrange( FDR ) 
```

# 10-Visualize top 25 genes on heatmap
```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=6}
topgenes <- rownames(pseudo.sig)[1:50]

scater::plotHeatmap(sce.sling, 
    order_columns_by='Pseudotime', 
    colour_columns_by="Concise_new1", 
    features = topgenes,
    center=TRUE,
    scale=TRUE)
```


# 11- GENES CHANGING ALONG THE TRAJECTORIES


## 11a- EXPRESSION OF TEK w/ individual data points


```{r, message=FALSE, warning=FALSE, echo=FALSE}

# Example for one gene (e.g., topgenes[1])
gene <- topgenes[39]
df <- data.frame(
  pseudotime = sce.sling$Pseudotime,
  expression = logcounts(sce.sling)[gene, ],
  cluster = sce.sling$Concise_new1
)

ggplot2::ggplot(df, aes(x = pseudotime, y = expression, color = cluster)) +
  ggplot2::geom_point(alpha = 0.5) +
  ggplot2::geom_smooth(method = "loess", se = FALSE) +
  ggplot2::labs(title = gene, y = "Log-normalized expression")

```

## 11b- EXPRESSION OF TEK w/o  individual data points

```{r, message=FALSE, warning=FALSE, echo=FALSE}
ggplot2::ggplot(df, aes(x = pseudotime, y = expression, color = cluster)) +
  ggplot2::geom_smooth(method = "loess", se = FALSE, size = 1.2) +
  ggplot2::labs(title = gene, y = "Log-normalized expression")

#topgenes

```


# 11c- Plotting Gene Expression of Tek (Tie2) on UMAP 

```{r, message=FALSE, warning=FALSE, echo=FALSE}


scater::plotReducedDim(sce.sling, dimred = "UMAP", colour_by = topgenes[39])


```



```{r, message=FALSE, warning=FALSE, echo=FALSE, eval= FALSE}

gh::gh("POST /user/repos", name = "SlingShot", private = TRUE) #making private repo


```



```{r, message=FALSE, warning=FALSE, echo=FALSE, eval= FALSE}

echo "# ðŸ“Š Slingshot Trajectory Inference and Visualization

This repository contains a complete R Markdown workflow for performing pseudotime trajectory inference using **Slingshot** and visualizing it with 2D/3D UMAP projections, pseudotime heatmaps, and gene expression trends.

---

## ðŸ“Œ Overview

This project demonstrates how to:

- Convert a Seurat object to a `SingleCellExperiment`
- Run trajectory inference with the `slingshot` package
- Visualize lineage paths using MST and smooth curves
- Project lineages onto UMAP (2D and 3D)
- Color by cluster and pseudotime using `plotly`
- Detect genes changing along trajectories using `TSCAN`
- Plot gene expression trends and heatmaps along pseudotime

---

## ðŸ“ File Structure

- `Sling.Rmd`: Main R Markdown file with all steps and visualizations
- `Sling.html`: Knitted HTML output with embedded plots
- `/figures/` (optional): Folder for saved figures (if not using `self_contained: true`)

---

## ðŸ§¬ Input

A preprocessed **Seurat object** with:

- PCA and UMAP dimensionality reductions
- Cluster labels (e.g., `Concise_new1`) stored in metadata

---

## ðŸš€ How to Use

1. Clone this repository or download the `.Rmd` and `.html` files.
2. Open `Sling.Rmd` in RStudio.
3. Ensure your Seurat object (e.g., `Endo_sickle`) is available in your environment.
4. Knit the file to generate the HTML report with embedded trajectory analysis.

---

## ðŸ“Š Outputs

- 2D UMAP plots with Slingshot MST and inferred trajectories
- Interactive 3D UMAP colored by cluster or pseudotime
- Pseudotime-ordered heatmaps of dynamic genes
- Smooth gene expression trends across pseudotime for selected markers

---

## ðŸ‘¤ Author

**Sedat Kacar**  
Postdoctoral Researcher  
Indiana University School of Medicine  
Division of Pulmonary, Critical Care, and Sleep Medicine

---

## ðŸ“œ License

Released under the MIT License. You may reuse, adapt, and redistribute with appropriate credit.
" >> README.md


```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval= FALSE}
commit <- "first commit for slingshot"

git init
git add .
git commit -m "first commit"

git branch -M main
git remote add origin https://github.com/sedatkacar56/SlingShot.git
git push -u origin main

```

```{r, message=FALSE, warning=FALSE, echo=FALSE, eval= FALSE}
commit <- "first commit for slingshot"

git init
git add .
git commit -m "first commit"

git branch -M main
git remote add origin https://github.com/sedatkacar56/SlingShot.git
git push -u origin main

```





